---
title: "Data Structures in R"
output: html_notebook
---

Load the required libraries:

```{r}
# For importing and exploring data
library(tidyverse)
# For reading in Excel files
library(readxl)
# For descriptive statistics
library(psych)
# For writing data to Excel
library(writexl)
```

Load the data:
```{r}
districts <- read.csv("project/data/districts.csv")
```
```{r}
star <- read_xlsx("project/data/star.xlsx")
```
Look at the data:
```{r}
View(star)
```
```{r}
glimpse(star)
```
```{r}
summary(star)
```
```{r}
describe(star)
```
```{r}
head(star)
```

## Column-wise operations

Select columns:
```{r}
select(star, tmathssk, treadssk, schidkn)
```

Drop columns:
```{r}
select(star, -tmathssk, -treadssk, -schidkn)
```

Pass all unwanted columns into a vector, then drop it:
```{r}
select(star, -c(tmathssk, treadssk, schidkn))
```

Select everything between two columns:
```{r}
star <- select(star, tmathssk:totexpk)
```

Create a column by combining reading and math scores:
```{r}
star <- mutate(star, combscores = treadssk + tmathssk)
head(star)
```

Rename a column:
```{r}
star <- rename(star, ttl_score = combscores)
head(star)
```

## Row-wise operations

Sorting rows:
```{r}
arrange(star, classk, treadssk)
```

To sort in descending order, use the `desc()` function:
```{r}
arrange(star, desc(classk), treadssk)
```

Filter rows:
```{r}
filter(star, classk == 'small.class')
```

Filter with conditions:
```{r}
filter(star, treadssk >= 500)
```

Filter with multiple conditions:
```{r}
filter(star, classk == 'small.class' & treadssk >= 500)
```

## Aggregating and joining data

```{r}
star_grouped <- group_by(star, classk)
head(star_grouped)
```

Let's summarize the data:
```{r}
summarize(star_grouped, avg_math = mean(tmathssk))
```

## Combining datasets

```{r}
districts <- read.csv("project/data/districts.csv")
head(districts)
```

```{r}
star <- read_xlsx("project/data/star.xlsx")
head(star)
```

We’ll provide the “base” table first (star) and then the “lookup” table (districts). The function will look for and return a match in districts for every record in star, or return NA if no match is found

```{r}
left_join(select(star, schidkn, tmathssk, treadssk), districts)
```

Group the data by class type
```{r}
star_grouped <- group_by(star, classk)
star_avg_reading <- summarize(star_grouped, avg_reading = mean(treadssk))
```
```{r}
star_avg_reading_sorted <- arrange(star_avg_reading, desc(avg_reading))
star_avg_reading_sorted
```

Let's recreate the previous steps using the pipe operator:
```{r}
star %>%
  group_by(classk) %>%
  summarize(avg_reading = mean(treadssk)) %>%
  arrange(desc(avg_reading))
```

Reshape the data using `tidyr`:
```{r}
star_pivot <- star %>%
                select(c(schidkn, treadssk, tmathssk)) %>%
                mutate(id = row_number())
```

```{r}
star_long <- star_pivot %>%
                pivot_longer(cols = c(tmathssk, treadssk),
                             values_to = "score", names_to = "test_type")
head(star_long)
```

```{r}
star_long <- star_long %>%
                mutate(test_type = recode(test_type,
                                          tmathssk = "math",
                                          treadssk = "reading"))

distinct(star_long, test_type)
```

Widen it back:
```{r}
star_wide <- star_long %>%
                pivot_wider(values_from = 'score', names_from = 'test_type')
head(star_wide)
```