---
title: "Workout 1"
output: github_document
---

## 1

Create a series of 10 elements, random integers from 70 to 100, representing scores on a monthly exam. Set the index to be the month names, starting in September and ending in June.

With this series, write code to answer the following questions:
- What is the student’s average test score for the entire year?
- What is the student’s average test score during the first half of the year (i.e., the first five months)?
- What is the student’s average test score during the second half of the year?
- Did the students improve their performance in the second half? If so, by how much?

```{r}
# Load necessary library
library(dplyr)
# Create a series of 10 random integers from 70 to 100
set.seed(123) # For reproducibility
months <- c("September", "October", "November", "December", "January", "February", "March", "April", "May", "June")
scores <- sample(70:100, 10)
# Create a data frame
scores_df <- data.frame(Month = months, Score = scores)

# Print the data frame
print(scores_df)
```

Calculate the average test score for the entire year, first half, and second half:
```{r}
# Calculate average scores
average_score <- mean(scores_df$Score)
first_half_average <- mean(scores_df$Score[1:5])
second_half_average <- mean(scores_df$Score[6:10])
difference <- second_half_average - first_half_average
# Print results
cat("Average score for the entire year:", average_score, "\n")
cat("Average score for the first half of the year:", first_half_average, "\n")
cat("Average score for the second half of the year:", second_half_average, "\n")
cat("Improvement in performance in the second half:", difference, "\n")
```

- In which month did the student perform best?
- What were the student’s five highest scores?
- Round the student’s scores to the nearest 10.

```{r}
# Find the month with the best performance
best_month <- scores_df$Month[which.max(scores_df$Score)]
best_score <- max(scores_df$Score)
# Find the five highest scores
highest_scores <- sort(scores_df$Score, decreasing = TRUE)[1:5]
# Round scores to the nearest 10
rounded_scores <- round(scores_df$Score, -1)
# Print results
cat("Best month:", best_month, "with score:", best_score, "\n")
cat("Five highest scores:", highest_scores, "\n")
cat("Rounded scores:", rounded_scores, "\n")
```

## 2

Generate 10 test scores between 40 and 60, again using an index starting with September and ending with June. Find the mean of the scores and add the difference between the mean and 80 to each of the scores.
```{r}
months <- c("September", "October", "November", "December", "January", "February", "March", "April", "May", "June")
# Create a series of 10 random integers from 40 to 60
scores2 <- sample(40:60, 10)
# Create a data frame
scores2_df <- data.frame(Month = months, Score = scores2)
# Print the data frame
print(scores2_df)
```

Calculate the mean of the scores and add the difference between the mean and 80 to each score:
```{r}
# Calculate mean score
mean_score2 <- mean(scores2_df$Score)
# Calculate difference from 80
difference2 <- 80 - mean_score2
# Add difference to each score
adjusted_scores2 <- scores2_df$Score + difference2
# Print results
cat("Mean score:", mean_score2, "\n")
cat("Difference from 80:", difference2, "\n")
cat("Adjusted scores:", adjusted_scores2, "\n")
```

## 3

Generate 10 random integers in the range 0 to 100

```{r}
# Create a series of 10 random integers from 0 to 100
scores3 <- sample(0:100, 10)
```
Create a series containing those numbers’ tens digits.
```{r}
# Create a series containing the tens digits
tens_digits <- scores3 %/% 10
# Print results
cat("Original scores:", scores3, "\n")
cat("Tens digits:", tens_digits, "\n")
```

Create a new series with 10 floating-point values between 0 and 1,000. Find the numbers whose integer component (i.e., ignoring any fractional part) are even.

```{r}
s <- runif(10, 0, 1000)
# Find the integer components that are even
even_integers <- s[as.integer(s) %% 2 == 0]
s
even_integers
```

## 4

Generate a series of 100,000 floats in a normal distribution with a mean of 0 and a standard deviation of 100.
```{r}
# Generate a series of 100,000 floats in a normal distribution
normal_series <- rnorm(100000, mean = 0, sd = 100)
# Print the first few values
head(normal_series)
```

Get the descriptive statistics of the series
```{r}
# Get descriptive statistics
summary(normal_series)
```

Replace the minimum value with five times the maximum value.

```{r}
# Replace the minimum value with five times the maximum value
min_value <- min(normal_series)
max_value <- max(normal_series)
normal_series[normal_series == min_value] <- 5 * max_value
```

Get the descriptive statistics of the series again to see the changes.
```{r}
summary(normal_series)
```

Demonstrate that 68%, 95%, and 99.7% of the values in s are indeed within one, two, and three standard distributions of the mean.
```{r}
# Calculate mean and standard deviation once
mean_value <- mean(normal_series)
sd_value <- sd(normal_series)

# Define a function to calculate proportion within n standard deviations
calc_proportion <- function(data, n_sd, mean_val, sd_val) {
  # Using mean() on a logical vector is more efficient than sum()/length()
  mean(data >= mean_val - n_sd * sd_val & data <= mean_val + n_sd * sd_val)
}

# Calculate proportions for different standard deviation ranges
sd_ranges <- c(1, 2, 3)
expected <- c(0.68, 0.95, 0.997)  # Theoretical values from empirical rule
proportions <- sapply(sd_ranges, function(n) calc_proportion(normal_series, n, mean_value, sd_value))

# Print results with expected values for comparison
for (i in seq_along(sd_ranges)) {
  cat(sprintf("Proportion within %d SD: %.4f (Expected: %.3f, Difference: %.4f)\n",
              sd_ranges[i], proportions[i], expected[i], proportions[i] - expected[i]))
}
```

Calculate the mean of numbers greater than s.mean(). Then calculate the mean of numbers less than s.mean(). Is the average of these two numbers the same as s.mean()?

```{r}
# Calculate mean of numbers greater than mean
mean_greater <- mean(normal_series[normal_series > mean_value])
# Calculate mean of numbers less than mean
mean_less <- mean(normal_series[normal_series < mean_value])
# Check if the average of these two numbers is the same as mean_value
is_same <- (mean_greater + mean_less) / 2 == mean_value
# Print results
cat("Mean of numbers greater than mean:", mean_greater, "\n")
cat("Mean of numbers less than mean:", mean_less, "\n")
cat("Is the average of these two numbers the same as mean_value?", is_same, "\n")
```

What is the mean of the numbers beyond three standard deviations?
```{r}
# Calculate mean of numbers beyond three standard deviations
mean_beyond_3sd <- mean(normal_series[normal_series < mean_value - 3 * sd_value | normal_series > mean_value + 3 * sd_value])
# Print result
cat("Mean of numbers beyond three standard deviations:", mean_beyond_3sd, "\n")
```

## 5

Create a series of 28 temperature readings, representing four seven-day weeks, randomly generated from a normal distribution with a mean of 20 and a standard deviation of 5, rounded to the nearest integer. Use the days of the week as the index.
```{r}
days <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
temps <- round(rnorm(28, mean = 20, sd = 5), 0)
temps_df <- data.frame(Day = rep(days, 4), Temp = temps)
temps_df
```

Mean of the temperatures on Mondays:

```{r}
mean_temp_mon <- round(mean(temps_df$Temp[temps_df$Day == "Mon"]), 0)
mean_temp_mon
```

Mean for all days of the week:
```{r}
mean_temp <- aggregate(temps_df$Temp, by = list(temps_df$Day), FUN = mean)
mean_temp <- round(mean_temp[, 2], 0)
mean_temp
```

Weekend mean:
```{r}
mean_temp_weekend <- round(mean(temps_df$Temp[temps_df$Day %in% c("Sat", "Sun")]), 0)
mean_temp_weekend
```

How many times is the change in temperature from the previous day greater than 2 degrees?
```{r}
temps_df$Temp_change <- c(NA, diff(temps_df$Temp))
count_greater_than_2 <- sum(temps_df$Temp_change > 2, na.rm = TRUE)
count_greater_than_2
```

What are the two most common temperatures in our data set, and how often does each appear?
```{r}
# Find the two most common temperatures
temp_freq <- table(temps_df$Temp)
most_common_temps <- sort(temp_freq, decreasing = TRUE)[1:2]
most_common_temps
```

## 6

Load the data from `project/data/taxi-passenger-count.csv`
```{r}
# Load necessary library
library(readr)
# Load the data
taxi_data <- read_csv("project/data/taxi-passenger-count.csv")
# Print the first few rows of the data
head(taxi_data)
```

Count how many times each passenger count appears in the data set. The data has only one unnamed column
```{r}
count_passenger <- table(taxi_data)
# Print the counts
print(count_passenger)
```

What percentage of taxi rides had only one passenger versus the (theoretical) maximum of six passengers.
```{r}
one_passenger_count <- sum(taxi_data == 1)
total_count <- nrow(taxi_data)
percentage_one_passenger <- (one_passenger_count / total_count) * 100
max_passenger_count <- sum(taxi_data == 6)
percentage_max_passenger <- (max_passenger_count / total_count) * 100
# Print results
cat("Percentage of rides with one passenger:", percentage_one_passenger, "%\n")
cat("Percentage of rides with maximum passengers (6):", percentage_max_passenger, "%\n")
```

Given these numbers, would more people benefit from smaller taxis that can take only one or two passengers or larger taxis that can take five or six passengers?

```{r}
# Calculate the total number of rides with 1 or 2 passengers
one_or_two_passenger_count <- sum(taxi_data == 1 | taxi_data == 2)
# Calculate the total number of rides with 5 or 6 passengers
five_or_six_passenger_count <- sum(taxi_data == 5 | taxi_data == 6)
# Calculate the percentage of rides with 1 or 2 passengers
percentage_one_or_two_passenger <- (one_or_two_passenger_count / total_count) * 100
# Calculate the percentage of rides with 5 or 6 passengers
percentage_five_or_six_passenger <- (five_or_six_passenger_count / total_count) * 100
# Print results
cat("Percentage of rides with 1 or 2 passengers:", percentage_one_or_two_passenger, "%\n")
cat("Percentage of rides with 5 or 6 passengers:", percentage_five_or_six_passenger, "%\n")
# Conclusion
if (percentage_one_or_two_passenger > percentage_five_or_six_passenger) {
  cat("More people would benefit from smaller taxis that can take only one or two passengers.\n")
} else {
  cat("More people would benefit from larger taxis that can take five or six passengers.\n")
}
```

## 7

```{r}
taxi_distance <- read_csv("project/data/taxi-distance.csv")
# Print the first few rows of the data
head(taxi_distance)
```

Create another series containing category names rather than numbers based on these criteria:
- Short, <= 2 miles
- Medium, > 2 miles but <= 10 miles
- Long, > 10 miles
```{r}
# First check the column names in the dataframe
names(taxi_distance)

# Then create categories using the correct column name
# Assuming the distance column is the first column
taxi_distance$Category <- cut(
  taxi_distance[[1]],  # Using the first column
  breaks = c(-Inf, 2, 10, Inf),
  labels = c("Short", "Medium", "Long")
)

# Alternative if the column has a different name
# Replace "actual_column_name" with the actual column name from names(taxi_distance)
# taxi_distance$Category <- cut(
#   taxi_distance$actual_column_name,
#   breaks = c(-Inf, 2, 10, Inf),
#   labels = c("Short", "Medium", "Long")
# )

# Show the first few rows of the result
head(taxi_distance)

# Count records in each category
table(taxi_distance$Category)
```

Visualize the distribution of the distances using a histogram.

```{r}
# Load necessary library
library(ggplot2)
# Create a histogram of the distances
ggplot(taxi_distance, aes(x = taxi_distance[[1]])) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Taxi Distances", x = "Distance (miles)", y = "Frequency") +
  theme_minimal()
```

How many short, medium, and long trips were there for trips that had only one passenger? Note that the data for passenger count and trip length is from the same data set, meaning the indexes are the same.

```{r}
# Load necessary library
library(readr)

# Load the data
passengers <- read_csv("project/data/taxi-passenger-count.csv")
distances <- read_csv("project/data/taxi-distance.csv")

# Create a new data frame with the columns we want
# Use first column from each dataframe regardless of column name
combined_data <- data.frame(
  Passenger_Count = unlist(passengers[,1]),
  Distance = unlist(distances[,1])
)

# Create categories based on distance
combined_data$Category <- cut(
  combined_data$Distance,
  breaks = c(-Inf, 2, 10, Inf),
  labels = c("Short", "Medium", "Long")
)

# Filter for trips with only one passenger
one_passenger_data <- combined_data[combined_data$Passenger_Count == 1, ]

# Count trips in each category
category_counts <- table(one_passenger_data$Category)

# Print the results
print(category_counts)
```
```{r}
# Calculate percentages
category_percentages <- prop.table(category_counts) * 100

# Create a visualization of the distribution
library(ggplot2)
ggplot(one_passenger_data, aes(x = Category)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Trip Distances for Single-Passenger Rides",
       x = "Distance Category",
       y = "Count")
```